Detailed structure of negatively charged membranes of phosphatidylserine and phosphatidylcholine at concentrations of calcium, sodium and potassium salts from molecular dynamics simulations with electronic polarization

### 2019-05-21 Manuscript/manuscript.tex
Edited section about calcium interactions

\begin{figure*}[!tbp] 
  \centering 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPS_nacl.pdf} 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPS_kcl.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_PC-PS_mix_PS} 
    The POPS head group order parameters, the P--N and C$_\beta$--C$_\gamma$ vector angles (stars)
    with respect to the membrane normal as a function of POPC content in a bilayer
    from ECC-lipids and Lipid17/Dang simulations with \ce{Na^+} (left) and \ce{K^+} (right) counterions.
    Experimental order parameter values and signs for pure POPS are from Ref. \citenum{NMRlipidsIV} and values
    for POPC:POPS (5:1) mixture from Ref. \citenum{roux90} 
    (only \ce{Na^+} counterions, but shown also in the right plots for \ce{K+}).
    % No translation in y-axis done in this plot:
    %The underestimated response of ECC-lipids with \ce{Na^+} counterions 
    %is likely due to a still slightly overestimated binding affinity of \ce{Na^+} to the phospholipids,
    %which is corroborated by the series with \ce{K^+} counterions (lower affinity than \ce{Na^+}),
    %where the response is closer to the experiment, which uses only \ce{Na^+} counterions. 
  }
\todo{J: I suggest to put this to SI. }
\end{figure*}

@@

%The headgroup order parameters of PC lipids decrease more upon addition of \ce{CaCl2}
%in POPC:POPS (5:1) mixtures than in pure PC bilayers due to the higher calcium binding
%affinity to the negatively charged lipid bilayers \cite{akutsu81,altenbach84,roux90,NMRlipidsIV}.
Our recent work demonstrates that the POPC headgroup order parameters measured
from POPC:POPS (5:1) mixture as a function of CaCl$_2$ concentration \cite{roux90}
can be used to evaluate the calcium binding affinity to lipid bilayers containing PS lipids \cite{NMRlipidsIV}.
The decrease of the PC headgroup order parameters in this mixture upon addition of \ce{CaCl2} 
were overestimated by the Lipid17/Dang simulations (also shown in Fig.~\ref{fig:delta_ordPar_CaCl_PCPS})
and all the other tested force fields, except CHARMM36 with the recently introduced NBfix for calcium \cite{kim16},
which underestimated the headgroup order parameter response.
In ECC-lipids simulations, the headgroup responses are in better agreement with experiments (Fig.~\ref{fig:delta_ordPar_CaCl_PCPS}),
indicating that the lower binding affinity than in Lipid17/Dang simulations is more realistic (Fig.~\ref{fig:cacl-dens_PCPS}).
%However, the steep decrease of the order parameters at low calcium concentrations,
%observed especially in negatively charged bilayers \cite{borle85,macdonald87,roux90},
%may not be fully captured by the ECC-lipids simulations.
The increased calcium binding affinity to bilayers containing PS lipids is illustrated in
ECC-lipids simulations by the increasing relative surface excess of calcium with respect to water from
$\Gamma^{w}_{Ca} = 0.06\mathrm{nm^{-2}}$ for pure POPC bilayer with 350 mM CaCl$_2$ \cite{melcr18}
to $\Gamma^{w}_{Ca} = 0.24\mathrm{nm^{-2}}$ for the POPC:POPS (5:1) bilayer with 409 mM CaCl$_2$.

@@

Experimentally measured headgroup order parameters of POPS exhibit a strong dependence
on low CaCl$_2$ concentrations with a rapid saturation below 100~mM \cite{roux90}.
Also these changes were overestimated by the Lipid17/Dang (also shown in Fig.~\ref{fig:delta_ordPar_CaCl_PCPS})
and other tested force fields in our recent work \cite{NMRlipidsIV}, 
but CHARMM36 with the NBfix correction for calcium
which underestimated the binding affinity.
In ECC-lipids simulations, the changes of the PS headgroup order parameters are not overestimated,
but the strong dependence on low concetrations of CaCl$_2$ is not fully reproduced (Fig.~\ref{fig:delta_ordPar_CaCl_PCPS}).
In addition to possibly suboptimal interactions between calcium ions and PS headgroup, the potential sources of this
discrepancy include the above observed slight overestimation of \ce{Na^+} counterions and
inaccurate structures of the lipid headgroup (Fig.~\ref{simVSexpNOions_POPS}).

In ECC-lipids simulations, the average orientation of P--N vectors of both POPC and POPS headgroups tilts more perpendicular to the membrane surface
by 11$^\circ$ and  3$^\circ$, respectively, upon addition of 620~mM CaCl$_2$ to POPC:POPS (5:1) mixture (Fig.~\ref{fig:delta_ordPar_CaCl_PCPS}),
suggesting that the PS headgroup orientation is less sensitive to the bound calcium than PC.
On the other hand, the average orientation of the C$_{\beta}$--C$_{\gamma}$ vector PS headgroup
tilts to the opposite direction by $11^\circ$ in the same system.
Such a largely diminished flexibility of the PS head group can be explained 
by the simultaneous binding of calcium to the \ce{COO^-} group in PS and phosphate group in PC. 
In Lipid17/Dang simulations, the changes in both headroup orientations and order parameters
are larger and less systematic for the PS headgroup, especially with lower concentrations (Fig.~\ref{fig:delta_ordPar_CaCl_PCPS}).
ECC-lipids simulations yield smaller error bars and shorter residence times (Fig.~\ref{fig:hist_residence_times}) in
than those typically observed in simulations with other force fields \cite{javanainen17,melcr18,NMRlipidsIV}.
This suggests that the ECC accelerates the equlibration of ions at lipid bilayer interface.

Our 1 $\mu$s simulations seems to be sufficiently long for the ECC-lipids simulations, because
90\% of the calcium residence times are shorter than $60\,\mathrm{ns}$ for pure POPC bilayer
and shorter than $200\,\mathrm{ns}$ for POPC:POPS (5:1) mixture, while the
longest observed residence times are $141\,\mathrm{ns}$ and $485\,\mathrm{ns}$, respectively (Fig.~\ref{fig:hist_residence_times}).
Interestingly, the calcium residence times are 3-4 times longer in POPC:POPS (5:1) mixture than in pure POPC.
In ECC-lipids simulations, on average 41\% of the total population of bound calcium cations is in contact with PS lipids
with 8\% bound only to them, even though the negatively charged membrane contains only 17\% of POPS (Table~\ref{tab:binding}).
Therefore, calcium ions bind more likely to negatively charged PS lipids than neutral PC lipid in a mixed bilayer, as expected.

@@

Recent analyzes of interaction sites between calcium and PS lipids combining
MD simulations with various experimental techniques have been 
controversial because the results strongly depend on force fields parameters \cite{melcrova16,valentine18,hallock18}.
In ECC-lipids simulation of POPC:POPS (5:1) mixture with 409 mM CaCl$_2$,
the calcium ions bind more likely 
to the carboxylate than to phosphate moiety of POPS headgroup (Fig.~\ref{fig:cacl_complexes}),
in line with the results without the NBfix corrections in CHARMM36 \cite{hallock18}.
However, in CHARMM36 simulations without the NBfix correction for calcium \cite{kim16},
calcium ions significantly overbind to PC lipids  \cite{catte16} and calcium ions coordinate
with even four distinct PS lipids \cite{hallock18}, while coordination with more than two PS lipids
was not observed in our ECC-lipids simulations (Fig.~\ref{fig:cacl_complexes}).
Furthermore, the POPS:POPC:Ca$^{2+}$ (7:3:7) mixture with relatively high
PS lipid fraction and calcium concentration was used in the previous work \cite{hallock18},
which may lead to the complexation of \ce{Ca^{2+}} cations and PS lipids that is
not observed with lower PS content used in ECC-lipids simulations in this work \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
%While POPC interacts with the calcium cations almost entirely through its phosphate group 
%in both neutral \cite{melcr18} and negatively charged membranes, 
%the inteaction with POPS lipids happens more through the \ce{COO^-} group than through the \ce{PO4} group. 
%Interactions of \ce{Ca^{2+}} with carbonyl groups are also present for both POPC and POPS,
%however, they are always accompanied by interactions with phosphate groups.

Significant coordination of calcium with carbonyls groups of PS and PC lipids observed in previous
simulations with the Berger force field \cite{melcrova16} %\cite{berger97,mukhopadhyay04}
was not observed in ECC-lipids simulations in here nor in the previous work with pure POPC \cite{melcr18}.
Overestimated coordination of cations with carbonyl groups is a potential source for
overestimated cation binding in Berger force fields \cite{catte16,NMRlipidsIV}. 
In CHARMM36 simulations with the NBfix, calcium mainly interacted with only
carboxylate of PS lipids  \cite{valentine18}, which is a potential source for the
underestimated calcium binding affinity to lipid bilayer with this force field \cite{NMRlipidsIV}.

@@

\begin{figure}[tb!] 
  \centering 
  \includegraphics[width=\figwidthsmall]{../img/bound_ca_populations.pdf} 
  \includegraphics[width=\figwidthsmall]{../img/stoichiometry_CaCl2_comparison_Ecc-lipids_PC-vs-PCPS.pdf} \\ 
  \caption{ \label{fig:cacl_complexes} 
  \textbf{Left:} Percentages of the population of bound Ca$^{2+}$ 
in exclusive contact with various lipid moieties 
in a POPC:POPS (5:1) bilayer with 409 mM CaCl$_2$. 
The bar representing total populations of bound \ce{Ca^{2+}} to PC, PS, or PC and PS (blue),
is in part covered by areas representing contributions of individual moieties exclusively (shades of orange and red). 
In the case of mixed simultaneous binding to PC and PS
we distinguish different moieties only in PS.
%i.e., PC can interact with any considered moiety. 
For comparison, in a pure POPC bilayer with 350 mM CaCl$_2$, 
67\% of bound \ce{Ca^{2+}} is exclusively in contact with \ce{PO_4} group. \cite{melcr18} 
    \textbf{Right:} Relative probabilities of \ce{Ca^{2+}} ions to coordinate with a certain number of lipids
    in pure POPC bilayer with 350 mM CaCl$_2$ and in POPC:POPS (5:1) mixture with 400 mM CaCl$_2$.  
    Analysis was done for both systems by considering all lipids (blue and violet) and
    for POPC:POPS (5:1) mixture also by considering POPC and POPS lipids separately (green and orange). 
    %All lipids were taken into account with the exception of the complexes in light green and orange, 
    %for which we counted only contacts with POPC resp POPS from the mixed 5\,PC:1\,PS negatively charged bilayer. 
    %The calculated probabilities of the calcium-lipid complexes also reflect only POPC (light green) resp POPS (orange). 
    % Probabilities were taken from simulations with comparable bulk concentrations of calcium around 250~mM. 
    Clusters of four or more lipids were not observed in either membrane.
    The threshold for counting coordinated lipids in a complex with \ce{Ca^{2+}} was set to $0.3\,\mathrm{nm}$, 
    as the distance between the cation and the oxygen atoms of the lipids. 
    Previously published simulation data \cite{melcr18} for pure POPC bilayers were taken directly from \cite{ECC-POPC_nacl_cacl2_files}. 
  }
\end{figure}

The advantage of the NMR order parameters used in this work is the direct connection between
measured and simulation data, which enables more unambiguous evaluation of the force field quality \cite{catte16,ollila16}
than more indirect comparison in previous studies using 2D infrared spectroscopy \cite{valentine18}, NMR chemical shifts and
rotational-echo double-resonance (REDOR) experiments \cite{hallock18} or fluorescent and vibrational sum frequency spectroscopy \cite{melcrova16}.
Because the ECC-lipids model gives the best agreement with the experimental headgroup order parameters
with various ion conctenrations among the available models, we believe that it gives the currently best available
interpretation of the experimental data. Athough the ECC-lipids model gives the most realistic picture of calcium binding to
lipid bilayers containing PS lipids with respect to NMR order parameters, it does not
correctly capture the rapid changes with small concentrations. Therefore, further
force field optimization is required for even more accurate MD simulations of cations
in the vicinity of negatively charged lipid bilayers.
% and also possibly from the approximate mean-field treatment of the electronic polarization.

@@

>>>>

@@

%The headgroup order parameters of PC lipids decrease more upon addition of \ce{CaCl2}
%in POPC:POPS (5:1) mixtures than in pure PC bilayers due to the higher calcium binding
%affinity to the negatively charged lipid bilayers \cite{akutsu81,altenbach84,roux90,NMRlipidsIV}.
Our recent work demonstrates that the POPC headgroup order parameters measured
from POPC:POPS (5:1) mixture as a function of CaCl$_2$ concentration 
can be used to evaluate the calcium binding affinity to lipid bilayers containing PS lipids \cite{roux90, NMRlipidsIV}.
The decrease of the PC headgroup order parameters in this mixture upon addition of \ce{CaCl2} 
were overestimated by the Lipid17/Dang simulations (also shown in Fig.~\ref{fig:delta_ordPar_CaCl_PCPS})
and all other tested force fields except CHARMM36 with the recently introduced NBfix for calcium \cite{kim16},
which underestimated the headgroup order parameter response.
In ECC-lipids simulations, the headgroup responses are in better agreement with experiments (Fig.~\ref{fig:delta_ordPar_CaCl_PCPS}),
indicating that the lower binding affinity than in Lipid17/Dang simulations is more realistic (Fig.~\ref{fig:cacl-dens_PCPS}).
%However, the steep decrease of the order parameters at low calcium concentrations,
%observed especially in negatively charged bilayers \cite{borle85,macdonald87,roux90},
%may not be fully captured by the ECC-lipids simulations.
The increased calcium binding affinity to bilayers containing PS lipids is illustrated in
ECC-lipids simulations by the increasing relative surface excess of calcium with respect to water from
%  previously we used 350mM conc, but I think it is better to compare to a higher concentration to support the whole argument bullet-proof
%  $\Gamma^{w}_{Ca} = 0.06\mathrm{nm^{-2}}$ ($\sim ?.?$ phospholipids per bound \ce{Ca^{2+}}) for pure POPC bilayer with 346~mM CaCl$_2$ \cite{melcr18, ECC-POPC_nacl_cacl2_files} 
   $\Gamma^{w}_{Ca} = 0.09\mathrm{nm^{-2}}$ ($\sim 5.4$ phospholipids per bound \ce{Ca^{2+}}) for pure POPC bilayer with 467~mM CaCl$_2$ \cite{ECC-POPC_nacl_cacl2_files} 
to $\Gamma^{w}_{Ca} = 0.24\mathrm{nm^{-2}}$ ($\sim 4.8$ phospholipids per bound \ce{Ca^{2+}}) for the POPC:POPS (5:1) bilayer with 409 mM CaCl$_2$.

@@

Experimentally measured headgroup order parameters of POPS exhibit a strong dependence
on low CaCl$_2$ concentrations with a rapid saturation below 100~mM \cite{roux90}.
Also these changes were overestimated by the Lipid17/Dang (also shown in Fig.~\ref{fig:delta_ordPar_CaCl_PCPS})
and other tested force fields in our recent work \cite{NMRlipidsIV}, 
but CHARMM36 with the NBfix correction for calcium
which underestimated the binding affinity.
In ECC-lipids simulations, the changes of the PS headgroup order parameters are not overestimated,
but the strong dependence on low concetrations of CaCl$_2$ is not fully reproduced (Fig.~\ref{fig:delta_ordPar_CaCl_PCPS}).
In addition to possibly suboptimal interactions between calcium ions and PS headgroup, the potential sources of this
discrepancy include the above observed slight overestimation of \ce{Na^+} counterions and
imperfect structures of the lipid headgroup (Fig.~\ref{simVSexpNOions_POPS}).

In ECC-lipids simulations, the average orientation of P--N vectors of both POPC and POPS headgroups tilts more perpendicular to the membrane surface
by 11$^\circ$ and  3$^\circ$, respectively, upon addition of 620~mM CaCl$_2$ to POPC:POPS (5:1) mixture (Fig.~\ref{fig:delta_ordPar_CaCl_PCPS})
suggesting that the PS headgroup orientation is less sensitive to the bound calcium than PC.
On the other hand, the average orientation of the C$_{\beta}$--C$_{\gamma}$ vector in PS headgroup
tilts to the opposite direction by $11^\circ$ in the same system.
In Lipid17/Dang simulations, the changes in both headroup orientations and order parameters
are larger and less systematic for the PS headgroup, especially with lower concentrations (Fig.~\ref{fig:delta_ordPar_CaCl_PCPS}).

ECC-lipids simulations yield smaller error bars and shorter residence times (Fig.~\ref{fig:hist_residence_times}) 
than those typically observed in simulations with other force fields \cite{javanainen17,melcr18,NMRlipidsIV}.
This suggests that the ECC accelerates the equlibration of ions at lipid bilayer interface.
Our 1 $\mu$s simulations seem to be sufficiently long for the ECC-lipids simulations, because
90\% of the calcium residence times are shorter than $60\,\mathrm{ns}$ for pure POPC bilayer
and shorter than $200\,\mathrm{ns}$ for POPC:POPS (5:1) mixture, while the
longest observed residence times are $141\,\mathrm{ns}$ and $485\,\mathrm{ns}$, respectively (Fig.~\ref{fig:hist_residence_times}).
Interestingly, the calcium residence times are 3-4 times longer in POPC:POPS (5:1) mixture than in pure POPC.

@@

Recent analyzes of interaction sites between calcium and PS lipids combining
MD simulations with various experimental techniques have been 
controversial because the results strongly depend on force fields parameters \cite{melcrova16,valentine18,hallock18}.
In ECC-lipids simulation of POPC:POPS (5:1) mixture with 409 mM CaCl$_2$,
on average 41\% of the total population of bound calcium cations is in contact with PS lipids
with 8\% bound exclusively only to them even though the negatively charged membrane contains only 17\% of POPS (Fig.~\ref{fig:cacl_complexes}).
%Therefore, calcium ions bind more likely to negatively charged PS lipids than neutral PC lipid in a mixed bilayer, as expected.
Also, calcium ions bind more likely 
to the carboxylate than to the phosphate moiety of POPS headgroup (Fig.~\ref{fig:cacl_complexes}),
in line with the results without the NBfix corrections in CHARMM36 \cite{hallock18}.
This can explain the diminished flexibility of the PS head group 
from simultaneous binding of \ce{Ca^{2+}} to the \ce{COO^-} group in PS and a phosphate group in PC 
as the most probable configuration (Fig.~\ref{fig:cacl_complexes}).

However, in CHARMM36 simulations without the NBfix correction for calcium \cite{kim16},
calcium ions significantly overbind to PC lipids  \cite{catte16} and calcium ions coordinate
with even four distinct PS lipids \cite{hallock18}, while coordination with more than two PS lipids
was very rare in our ECC-lipids simulations (Fig.~\ref{fig:cacl_complexes}).
Furthermore, the POPS:POPC:Ca$^{2+}$ (7:3:7) mixture with relatively high
PS lipid fraction and calcium concentration was used in the previous work \cite{hallock18},
which may lead to the complexation of \ce{Ca^{2+}} cations and PS lipids that is
not observed with lower PS content used in ECC-lipids simulations in this work \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
%While POPC interacts with the calcium cations almost entirely through its phosphate group 
%in both neutral \cite{melcr18} and negatively charged membranes, 
%the inteaction with POPS lipids happens more through the \ce{COO^-} group than through the \ce{PO4} group. 
%Interactions of \ce{Ca^{2+}} with carbonyl groups are also present for both POPC and POPS,
%however, they are always accompanied by interactions with phosphate groups.

Significant coordination of calcium with carbonyl groups of PS and PC lipids observed in previous
simulations with the Berger force field \cite{melcrova16} %\cite{berger97,mukhopadhyay04}
was not observed in ECC-lipids simulations in here nor in the previous work with pure POPC \cite{melcr18}.
Overestimated coordination of cations with carbonyl groups is a potential source for
the overestimated cation binding in Berger force field \cite{catte16,NMRlipidsIV}. 
In CHARMM36 simulations with the NBfix, calcium mainly interacted with only
carboxylate of PS lipids  \cite{valentine18}, which is a potential source for the
underestimated calcium binding affinity to lipid bilayer with this force field \cite{NMRlipidsIV}.

@@

\begin{figure}[tb!] 
  \centering 
  \includegraphics[width=\figwidthsmall]{../img/bound_ca_populations.pdf} 
  \includegraphics[width=\figwidthsmall]{../img/stoichiometry_CaCl2_comparison_Ecc-lipids_PC-vs-PCPS.pdf} \\ 
  \caption{ \label{fig:cacl_complexes} 
  \textbf{Left:} Percentages of the population of bound Ca$^{2+}$ 
in exclusive contact with the given oxygen moiety
in a POPC:POPS (5:1) bilayer with 409 mM CaCl$_2$. 
The areas represent populations of bound \ce{Ca^{2+}} 
only to the given moieties exclusively (shades of orange and red)
or their combinations (CHANGE:blue)
separately for PC, PS, or PC and PS simultaneously. 
The areas sum up to total populations at the respective lipids. 
In the case of mixed simultaneous binding to PC and PS
we distinguish different moieties only in PS.
%i.e., PC can interact with any considered moiety. 
For comparison, in a pure POPC bilayer with 350 mM CaCl$_2$, 
67\% of bound \ce{Ca^{2+}} is exclusively in contact with \ce{PO_4} group. \cite{melcr18} 
    \textbf{Right:} Relative probabilities of \ce{Ca^{2+}} ions to coordinate with a certain number of lipids
    in pure POPC bilayer with 350 mM CaCl$_2$ and in POPC:POPS (5:1) mixture with 400 mM CaCl$_2$.  
    Analysis was done for both systems by considering all lipids (blue and violet) and
    for POPC:POPS (5:1) mixture also by considering POPC and POPS lipids separately (green and orange). 
    %All lipids were taken into account with the exception of the complexes in light green and orange, 
    %for which we counted only contacts with POPC resp POPS from the mixed 5\,PC:1\,PS negatively charged bilayer. 
    %The calculated probabilities of the calcium-lipid complexes also reflect only POPC (light green) resp POPS (orange). 
    % Probabilities were taken from simulations with comparable bulk concentrations of calcium around 250~mM. 
    Clusters of four or more lipids were not observed in either membrane.
    The threshold for counting coordinated lipids in a complex with \ce{Ca^{2+}} was set to $0.3\,\mathrm{nm}$, 
    as the distance between the cation and the oxygen atoms of the lipids. 
    Previously published simulation data \cite{melcr18} for pure POPC bilayers were taken directly from \cite{ECC-POPC_nacl_cacl2_files}. 
  }
\todo{J: ad bar plots -- change total to combinations as described in caption, 
add simultaneously to "PC and PS", 
add ticks to the right side of the plot and
change the colour of "combinations" from blue to any other so that it is different from the legend in the other plot. }
\end{figure}

@@

The advantage of the NMR order parameters used in this work is the direct connection between
measured and simulation data, which enables more unambiguous evaluation of the force field quality \cite{catte16,ollila16}
than more indirect comparison in previous studies using 2D infrared spectroscopy \cite{valentine18}, NMR chemical shifts and
rotational-echo double-resonance (REDOR) experiments \cite{hallock18} or fluorescent and vibrational sum frequency spectroscopy \cite{melcrova16}.
Because the ECC-lipids model gives the best agreement with the experimental headgroup order parameters
with various ion conctenrations among the available models, we believe that it also gives the currently best available
interpretation of the experimental data. Athough the ECC-lipids model gives the most realistic picture of calcium binding to
lipid bilayers containing PS lipids with respect to NMR order parameters, it does not
correctly capture the rapid changes with small concentrations. Therefore, further
force field optimization is required for even more accurate MD simulations of cations
in the vicinity of negatively charged lipid bilayers.
% and also possibly from the approximate mean-field treatment of the electronic polarization.
